---
title: "class05: data vis w/ ggplot"
author: "Annmarie (PID:16442048)"
format: html
---

# Graphics systems in R

There are many graphics systems in R for making plots and figures.

We have already played a little with **"Base R"** graphics and the `plot()` function.

Today we sill start learning about a popular graphics package called `ggplot2`().

Thus is an add on package - ie. we need to install it. I install it (like I install any package) with the `install.packages()` function.

```{r}
plot(cars)
```

Before I can use the functions from a package I have to load the package from my "library". We use the `library(gglot2)` command to look it up

```{r}
library(ggplot2)
ggplot(cars)
```

Every ggplot is made up of at least 3 things: - data (the numbers etc. that will go into your plot) - aes ( how the columns of data map to the plot aesthetics) - geoms (how the plot actually looks, points, bars, lines, etc.)

```{r}
ggplot(cars)+
  aes(x=speed, y=dist)+
  geom_point()

```

For simple plots ggplot is more verbose - it takes more to code - than base R plot.

Add some more layers to our ggplot:

```{r}
ggplot(cars)+
  aes(x=speed, y=dist)+
  geom_point()+
  geom_smooth()+
  labs(title="Stopping distance of old cars",
subtitle = "A silly example plot")+
  theme_bw()

```

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

```{r}
nrow(genes)
```

```{r}
colnames(genes)
```

```{r}
ncol(genes)
```

```{r}
table(genes$State)
```

```{r}
round(table(genes$State)/nrow(genes)*100,2)
```

```{r}
p <- ggplot(genes)+
  aes(x=Condition1, y=Condition2, col=State)+
  geom_point()
p
```

```{r}
p + scale_colour_manual( values=c("blue","gray","red"))+
  labs(subtitle = "Gene Expression Changes Upon Drug Treatment")
```
```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```
```{r}
# install.packages("dplyr")  ## un-comment to install if needed
library(dplyr)

gapminder_2007 <- gapminder %>% filter(year==2007)
```

```{r}
ggplot(gapminder_2007)+
  aes(x=gdpPercap, y=lifeExp, color=continent, size=pop)+
  geom_point(alpha=0.5)
```

```{r}
ggplot(gapminder_2007)+
  aes(x=gdpPercap, y= lifeExp, size = pop)+
  geom_point(alpha=0.5)
```

```{r}
ggplot(gapminder_2007)+
  aes(x=gdpPercap, y= lifeExp, size = pop)+
  geom_point(alpha=0.5)+
  scale_size_area(max_size=10)
```
```{r}
gapminder_1957 <- gapminder %>% filter(year==1957)
```

```{r}
colnames(gapminder_1957)

```

```{r}
ggplot(gapminder_1957)+
  aes(x=gdpPercap, y=lifeExp)+
  geom_point()
```

```{r}
ggplot(gapminder_1957)+
  aes(x=gdpPercap, y=lifeExp, color=continent, size=pop)+
  geom_point()
```
```{r}
gapminder_both<- gapminder %>% filter (year==1957 | year==2007)
```
```{r}
ggplot(gapminder_both)+
geom_point(aes(x = gdpPercap, y = lifeExp, color=continent,
                 size = pop), alpha=0.7) + 
  scale_size_area(max_size = 10) +
  facet_wrap(~year)
```

